### 2023-05-08 response-to-audit-report.md

## Overview of Isabelle Changes

The audit report's main and high-severity concerns for the Isabelle Proofs were:

- 2.1.6 Missing description of Merkleization
- 2.1.4 Inaccurate formulation of Money preservation

We didn't mention the merkleization in the specification document as we thought of it as an implementation detail. This does lessen the applicability of the Isabelle proofs to the different implementations (marlowe-cardano, purescript-marlowe, etc). To check that an implementation is compliant with the specification we use the `spec-test` tool, which uses property based tests to check that a custom implementation behaves equally to the Isabelle implementation. Without the proper formalization, the tool can only test for the non-merkleized cases. In practice this gives us confidence that the merkleized interpreter works the same for non-merkleized cases. To test the correctness of merkleized cases, the marlowe-cardano repository has specific tests. The Merkleization formalization will be introduced in a future version of the specification, the initial formulation started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).

@@

Although this `Eq` instance is not used by the Marlowe validators (so it does not have implications for validator security), the equality test has been fixed because it is a liability for off-chain code and for potential future versions of validators that might perform such an equality test. Commit [84d65a70](https://github.com/input-output-hk/marlowe-cardano/commit/84d65a70c29320fab501ad759bbc71855d1b63fc) fixes this and commit [04805a39](https://github.com/input-output-hk/marlowe-cardano/commit/04805a393c0815fe0f2a1a6d7f76d9867ffe0c14) hardens all of the other `Eq` instances to prevent similar problems re-arising if sum types are augmented in the future.

### 2.1.4 Inaccurate formulation of Money preservation *(Severity: High)*

> **File `specification-v3-rc1.pdf`, Section `3.1 Money preservation`, page *29***
>
> As the property stands, it is permitted to make deposits in one currency and return payments in a different currency. As long as the sums of the amounts match, the equality is satisfied. Yet it is unlikely that the participants of the contract would agree that money has been preserved.
>
> Money preservation is a property stated with an equality. The left hand side is the sum of the deposits done by a list of transactions. The right hand side of the equality is the sum of all the payments done in the same list of transactions. Each sum, in turn, is represented as a single integer which aggregates the amounts of the various payments and deposits, irrespective of what currencies correspond to these amounts.

To address this, a new `MultiAssets.thy` theory was created. This theory defines a new type that ties together the number of tokens and the type of token, and it implements some type classes to be able to work with normal addition and substraction. Also, the MoneyPreservation was refactored into AssetsPreservation, addressing issue 2.1.4 among others.

These changes can be found in [PR-161](https://github.com/input-output-hk/marlowe/pull/161)

@@

This type was properly documented in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

### 2.2.7 Code snippets switch languages *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.1 Compute Transaction`, page *15***
>
> The specification changes from using Isabelle to using Haskell henceforth. Making the reader aware of the criteria for the language change would help maintaining the document.

The cause of this problem was that initially we had the Isabelle code separated from the documentation generation and we couldn't find a way to display the Isabelle code, so we showed the exported Haskell instead. We agree that it is confusing so we refactored to a literate programming style in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

@@

The way we document evalValue was modified in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

### 2.2.15 Missing parentheses in `div` specification *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, page *25***
>
> On page 25 formula $$c \neq 0 \Rightarrow c \mathbin{∗} a \mathbin{\mathrm{div}} (c \mathbin{∗} b) = a \mathbin{\mathrm{div}} b$$ needs additional parentheses around the term $c \mathbin{∗} a$, otherwise it can be parsed as $$c \neq 0 \Rightarrow c \mathbin{∗} (a \mathbin{\mathrm{div}} (c \mathbin{∗} b)) = a \mathbin{\mathrm{div}} b$$ which does not hold (Counter-example: $c=2, a=3, b=2$). The lemma `divMultiply` in the file `Semantics.thy` does use extra parentheses around $c \mathbin{∗} a$.

This lemma was removed from the PDF specification and replaced with lemmas that use `evalValue` instead.

@@

The way we document evalValue was modified in [PR-182](https://github.com/input-output-hk/marlowe/pull/182)

### 2.2.18 Missing `evalValue` lemmas in specification *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, pages *23--26***
>
> Not all lemmas about `evalValue` are listed in the specification. The absent lemmas include `evalDoubleNegValue`, `evalMulValue`, `evalSubValue`, and all division lemmas.

This section was rewritten and only the Division lemmas are now present as they are the most relevant ones.

@@

### 2.2.19 Typo in **Use Value** case of `evalValue` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.2.10 Evaluate Value`, page *26***
>
> The **Use Value** case mentions `TimeIntervalEnd` instead of `UseValue`.

This section was [rewritten](https://github.com/input-output-hk/marlowe/pull/182) and the issue was fixed

@@

The function `playTrace` was documented in [PR-182](https://github.com/input-output-hk/marlowe/pull/182) in section 2.2.7.

### 2.2.21 Type parameter discrepancy in `playTrace` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3 Marlowe Guarantees`, page *28***
>
> The first parameter of `playTrace` in the specification is `int`, while it is `POSIXTime` in the code. Even though the latter is an alias for the former, it is beneficial to use the `POSIXTime` name both for consistency and readability.

This happened because we were using a type alias, which gets fully expanded when being referenced by the documentation. [PR-182](https://github.com/input-output-hk/marlowe/pull/182) addresses this by making the whole theory in literal programming mode.

@@

The proposed definition is simpler and it is likely to be implemented in the future.

### 2.2.24 Discrepancy with Isabelle code for `allAccountsPositive` *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Sections `3.3 Positive Accounts`, page *30***
>
> The `allAccountsPositive` function is defined differently in the specification and in the Isabelle code, although both definitions show the same behavior. These definitions need to be consolidated.

The definition in the specification is the exported Haskell code of the Isabelle code. This is a limitation that Isabelle has when importing code from other theories. A possible solution is to change the Accounts positive theory to be literal programming.

@@

This was addressed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

### 2.2.27 Function `isClosedAndEmpty` is unexplained *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `3.6.2 All Contracts Have a Maximum Time`, page *32***
>
> The function `isClosedAndEmpty` needs to be explained.

In [PR-186](https://github.com/input-output-hk/marlowe/pull/186) we added a small explanation; in the future, we'll probably expand on this by refactoring the guarantees as literate programming.

@@

## 2.3 Lemmas and proofs

### 2.3.1 Unnecessarily large proofs *(Severity: Medium)*

> **Several Isabelle files, several lemmas**
>
> Some Isabelle proofs are written with long apply-scripts, where Isar would document the proof better. Proofs could also be split using more auxiliary lemmas.
>
> As the proofs stand, it is hard to figure out why a proof step fails, after changes elsewhere required a proof to be updated. Since the newly-failing proof step was designed with specific goals in mind, and changes in the code may lead to it facing a different set of goals, the maintainer might need to reconstruct the whole structure of the proof from an older version to infer state that Isabelle produces at each step.
>
> What Isar brings is making the intention of the author explicit at every step of the proof. This helps the maintainer of the proofs and fixes the concerns mentioned above.
>
>  will likely have to update the proofs. We conjecture that it will happen at least every time they target a new platform. In the case of Cardano, they need to extend the semantics to explain Merkleization. Another action that would make long proofs easier to understand is to split them using more auxiliary lemmas, thus feeding the information to the reader in smaller chunks.
>
> Some examples of large proofs:
> -   in `MoneyPreservation.thy`, lemmas `reduceContractStep_preserves_money` and\ `reductionLoop_preserves_money`
> -   in `SingleInputTransactions.thy`, lemmas `applyAllInputsPrefix_aux`,\ `computeTransactionIterative`, and `computeTransactionStepEquivalence_error`

We agree with this observation. New lemmas and theories are proven with this recommendation in mind, and some theories were refactored (like the Asset preservation). It will take some time until all theories are in this format.

@@

Same response as 2.3.1

### 2.3.5 Involved proof of `insert_valid` *(Severity: Low)*

> **File `MList.thy`, theorem `insert_valid`, line *66***
>
> The proof of `insert_valid` sprouts three other lemmas of difficult characterization: `insert_valid_aux`, `insert_valid_aux2`, and `insert_valid_aux3`. These lemmas make assumptions with implications that get in the way of understanding them in isolation.
>
> An alternative to make the proof pieces more reusable is to use instead the following set of lemmas, which also offers insight on how function `insert` interacts with predicates `sorted` and `distinct`:
>
> ``` {.text escapeinside="||" mathescape="true"}
>   lemma insert_set:
>     "set (map fst (insert a b xs)) = set (map fst xs) |$\cup$| { a }"
>
>   lemma insert_sorted:
>     "List.sorted (map fst c) |$\Longrightarrow$| List.sorted (map fst (insert a b c))"
>
>   lemma insert_distinct :
>     "List.distinct (map fst c)
>      |$\Longrightarrow$|
>      List.sorted (map fst c)
>      |$\Longrightarrow$|
>      List.distinct (map fst (MList.insert a b c))"
> ```
>
> which then can be combined in the proof of `insert_valid` as follows:
>
> ``` {.text escapeinside="||" mathescape="true"}
>   theorem insert_valid2 : "valid_map c |$\Longrightarrow$| valid_map (MList.insert a b c)"
>     using insert_sorted[of c a b] insert_distinct[of c a b] by fastforce
> ```
>
> The proofs of the lemmas can be found in […].

We agree that the proposed proof is easier, but the lemma will become obsolete once we replace the custom MList with
`HOL-Library/finiteMap`, which addresses other problems found in the report.

@@

The valid and positive preservation is needed in lower-level functions so that higher-level functions can use that fact when proving other properties. This makes `playTrace` not needing that proof as it is the highest-level function. The theorem  could be useful on its own and might get added in the future.

### 2.3.12 Unconcise goal in `reduceContractStepPayIsQuiescent` *(Severity: Low)*

> **File `QuiescentResult.thy`, lemma `reduceContractStepPayIsQuiescent`, line *8***
>
> This lemma does not express its goal concisely, as it makes no mention of `reduceContractStep` in the formulation. Changing the first assumption to $\texttt{reduceContractStep}\ \mathit{env}\ \mathit{sta}\ (\texttt{Pay}\ \mathit{x21}\ \mathit{x22}\ \mathit{tok}\ \mathit{x23}\ \mathit{x24})$ makes more explicit in which contexts this lemma can be useful. Modifying this assumption requires an additional `apply simp` to be added to the proof (before line 30) for the lemma to go through. Further, an additional `apply simp` will need to be added in lemmas `reduceContractStepIsQuiescent` (before line 44) and `timedOutReduce_only_quiescent_in_close` (`Timeout.thy`, before line 128) as well.

We agree on this observation and it will be addressed in a future refactor of the theory file.

@@

These issues were resolved in [PR-168](https://github.com/input-output-hk/marlowe/pull/168) and [PR-169](https://github.com/input-output-hk/marlowe/pull/169)

### 2.3.23 Inconsistent style when applying constructor *(Severity: Low)*

> **File `SingleInputTransactions.thy`, lemmas `beforeApplyAllLoopIsUseless, fixIntervalOnlySummary`, lines *275, 398***
>
> The lines mentioned in these lemmas display the resulting constructor before the function application, which differs from the general style in the rest of the codebase.

These observations regarding readability will be addressed when we refactor the theory.

@@

This observation regarding readability will be addressed when we refactor the theory.

### 2.3.28 `playTrace_only_accepts_maxTransactionsInitialState` not written as `theorem` *(Severity: Low)*

> **File `TransactionBound.thy`, lemma `playTrace_only_accepts_maxTransactionsInitialState`, line *316***
>
> This lemma seems like the main result of this file. Assuming it is an important result, we recommend writing it as a `theorem` rather than a `lemma`.

We agree on the observation, and will likely promote it to theorem in the future.

@@

We removed the unnecessary assumptions in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).

@@

## 2.4 Isabelle implementation

### 2.4.1 Variable shadowing in `applyAllLoop` *(Severity: Medium)*

> **File `Semantics.thy`, function `applyAllLoop`, line *575***
>
> The $\texttt{cont}$ variable introduced by the pattern match shadows another $\texttt{cont}$ variable, coming from the pattern match of an outer case expression, making the function harder to follow while also making it more error-prone to future changes.

The variable was renamed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168).

@@

### 2.4.2 Undescriptive name `moneyInPayment` *(Severity: Low)*

> **File `MoneyPreservation.thy`, function `moneyInPayment`, line *5***
>
> The name of the function can be more precise. Perhaps `moneyInPaymentToParty` or `moneyInExternalPayment` would work.

The theory was refactored into AssetsPreservation and the function is now called `assetsInExternalPayment`.

@@

The typo was fixed in [PR-168](https://github.com/input-output-hk/marlowe/pull/168)

### 2.4.5 Unclear need for multiple formulations for positive accounts *(Severity: Low)*

> **File `PositiveAccounts.thy`, `Throughout`**
>
> It is unclear what the use is for multiple formulations (and lemmas about) positive accounts. The first formulation (with the theorems `playTraceAux_gtZero` and `playTrace_gtZero`) is not used in any other modules but the alternative formulation is used instead. If both formulations are relevant, then it should be explained why.

We agree that it is confusing, and it will be refactored in the future.

@@

We agree with the observation and it will be addressed in a future refactor.

### 2.4.10 Misleading function names *(Severity: Low)*

> **File `SingleInputTransactions.thy`, function `inputsToTransactions`, line *9***
>
> This function name is not very descriptive of its meaning. It takes a transaction (both a time interval and a list of inputs) and returns a list of transactions at the same interval containing a single input each. A name like `splitTransactionIntoSingleInputTransactions` would convey better what the input and the output are.
>
> Moreover, the code would be cleaner if the function takes a single argument of type `Transaction`, instead of asking the caller to rip apart its fields.
>
> **File `SingleInputTransactions.thy`, function `traceListToSingleInput`, line *18***
>
> This function name is not descriptive of what it does. Perhaps a more telling name could be `splitTransactionsIntoSingleInputTransactions`.
>
> **File `SingleInputTransactions.thy`, function `isSingleInput`, line *1222***
>
> This function should be renamed or repurposed. If renamed, `allAreSingleInput` more accurately reflects the meaning of the function. If repurposed, it should check that a single transaction has a single input, and `all isSingleInput` can be used to express the current behavior.

We agree on the observations, for the moment some simple refactoring and comments were added as part of
[PR-180](https://github.com/input-output-hk/marlowe/pull/180/files#diff-638dc578b752590b5f2bace45f1e1ae9e522351f62056e6f92aa272424657b03). Further refactor and optional renaming might happen in the future.

@@

### 2.4.11 Unused parameter in `maxTransactionCaseList` *(Severity: Low)*

> **File `TransactionBound.thy`, function `maxTransactionCaseList`, line *16***
>
> This function has a parameter of type `State` that is completely unused and can be removed.

We agree on this observation and it will addressed in a future refactoring of the theory file.

### 2023-05-09 response-to-audit-report.md

# Response to Audit Report for the Marlowe Specification and for Marlowe on Cardano

@@

## Executive Summary

Marlowe consists of a domain-specific language (DSL) for the financial industry and for peer-to-peer financial contracts, with a reference semantics written in the Isabelle theorem-proving language and an implementation for the Cardano blockchain written in the Haskell/Plutus programming language. Early in 2023 the Marlowe specifications, on-chain implementation, and validator-test suite were audited by Tweag's High Assurance Software Group: their report is available online and it categorizes findings as either high, medium, or low severity. The present document summarizes the subsequent changes, made in response to the audit's findings, to the Marlowe specifications, its implementation as a Plutus smart-contract validator, and the property-based tests for that validator. All of the high-severity findings have been remedied or mitigated through changes to specifications, proofs, or implementation. Nearly all of the medium- and low-severity ones have also been addressed; justification for not addressing the remainder has been provided.

The high-severity findings related to handling of negative deposits, prevention of "double satisfaction", enforcement of state invariants, an implementation difference between the formal specification versus the Plutus implementation, and the Isabelle proof of money preservation. Marlowe's Plutus validator has been altered to resolve these first three findings so that it correctly prevents negative deposits from altering the balance of internal accounts, it only allows other Plutus scripts to run during transactions that do not make Marlowe payments, and it rigorously enforces invariants for initial and final states. The fourth finding regarding the Plutus implementation was mitigated via code analysis and property-based testing. The fifth finding about the money-preservation proof was remedied by revising the Isabelle proof.

@@

The medium-severity findings typically suggested improvements to the Isabelle proofs, addition of more extensive property-based testing, correction of infelicities in the implementation, or safe-coding improvements. The low-severity findings generally centered on the usefulness of code comments, naming of variable bindings, omissions in specifications, and correction of typographical errors. All of the findings that affect the clarity of the specifications, the coverage of testing, or the robustness of implementation were made. Numerous additional property-based tests were added to the test suite as a result of audit findings.

The changes outlined here have not been re-audited. The alterations to Marlowe's Plutus validator (detailed in the Appendix) have been kept to the minimum needed to address the audit findings, but avoiding extensive changes that would might have lessened the relevance of the audit.

Finally, please note that the audit's scope included the Marlowe language and its Plutus interpreter, but not individual Marlowe contracts or off-chain code. Although the Marlowe language and its implementation on Cardano aim to make smart contracts safer, it is nevertheless still possible to design individual instances of Marlowe contracts that exhibit unwanted behavior. Such undesirable behavior may result either from poor contract design or lack of testing and analysis of the contract. [A best-practices guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/best-practices.md) and [a security guide](https://github.com/input-output-hk/marlowe-cardano/blob/main/marlowe/security.md) provide advice for the safe and secure design and testing of individual Marlowe contracts, which can themselves be subject to audit or certification.

@@

## Acknowledgement

We thank the High Assurance Software Group at Tweag for their diligent and expert analyses of the Marlowe specifications, proofs, implementations, and tests and their clear and thorough communication of their findings.

## Overview of Isabelle Changes

The audit report's main and high-severity concerns for the Isabelle Proofs were:

- 2.1.6 Missing description of Merkleization
- 2.1.4 Inaccurate formulation of Money preservation

We did not mention the merkleization in the specification document as it was considered an implementation detail. This does lessen the applicability of the Isabelle proofs to the different implementations (Marlowe on Cardano, Marlowe in PureScript, etc). To check that an implementation is compliant with the specification we use the `spec-test` tool, which uses property based tests to check that a custom implementation behaves equally to the Isabelle implementation. Without the proper formalization, the tool can only test for the non-merkleized cases. In practice this gives us confidence that the merkleized interpreter works the same for non-merkleized cases. To test the correctness of merkleized cases, the marlowe-cardano repository has specific tests. The Merkleization formalization will be introduced in a future version of the specification, the initial formulation started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).

The Money preservation theory had only taken into account the amount of assets and not the type. This meant that the proof was not considering a case where a contract could receive `20 ada`,  `15 djed`, and pay back `20 djed` and `15 ada`. We added a new `MultiAssets` type and refactored the Isabelle code (without modifying the interpreter) to prove that the Assets are preserved.

Other medium and low priorities findings regarding documentation and legibility of the specification were also addressed.

@@

1. Incorrect handling of negative deposits
    - "2.1.1 Negative deposits allow stealing funds"
2. Incomplete prevention of double satisfaction
    - "2.1.2 Contracts vulnerable to double satisfaction attacks"
3. Lax enforcement of invariants in Marlowe state
    - "2.1.7 Positive balances are not checked for the output state"
    - "2.1.8 Non-validated Marlowe states"
    - "2.1.9 Total balance of ending state uncomputed"
    - "2.1.10 Unchecked ending balance"
4. Difference semantics of association lists in Isabelle and Plutus
    - "2.1.12 Different insertion functions used in Isabelle and Haskell code"

Incorrect handling of negative deposits (item 1 above) was mitigated by the addition of a guard against this behavior in Marlowe's Plutus validator so that its behavior is consistent with Marlowe semantics. Property-based tests were also added to check the correctness of the mitigation.

Although the Marlowe validator had prevented double-satisfaction among multiple copies of the Marlowe validator script running in the same transaction, it did not prevent double-satisfaction in cases where the Marlowe validator ran alongside another Plutus validator in the same transaction (item 2 above). Double satisfaction is now prevented by enforcing that the Marlowe validator is the only Plutus script that runs during transactions that make payments to parties. This allows Marlowe contracts to coordinate with other Plutus scripts, but only under conditions where double satisfaction is impossible. Once again, property-based tests were added to check the correctness of this mitigation.

The Marlowe validator had made optimistic assumptions about its own correct operation and hence did not check certain invariants in order to reduce Plutus execution costs (item 3 above). The Marlowe semantics validator has now been thoroughly hardened against corruption of initial or final state so that it ensures that the three state-invariants of positive accounts, non-duplication of state entries (accounts, choices, and bound values), and a total internal value matching the script UTxO's value. This hardening was completed without unduly increasing the size (in bytes) of the validator or its Plutus execution cost: such was verified with simulated and on-chain measurement of execution costs for a library of real-life Marlowe contracts. Notably, the enforcement of the final invariants protects against funds becoming permanently locked in a Marlowe contract due to a garbled state: even in cases where the implementation of Marlowe semantics or parts of the Plutus validator were flawed, at least some of the execution paths of a Marlowe contract might remain viable, so funds could be released even under nominally impossible but catastrophic difficulties. The property-based test suite was significantly enhanced to check double-satisfaction situations.

@@

Mitigation of the difference between association lists (item 4 above) in Isabelle (`MList`) and Plutus (`AssocMap`) was handled by the aforementioned enforcement of invariants, by line-by-line code inspection and annotation, and by thoroughly enhanced property-based testing. Work is in progress on a formal proof of the equivalence of `MList` and `AssocMap` under pre-conditions that hold within Marlowe semantics. Note that porting Isabelle's `MList` implementation to Plutus would have enlarged the Marlowe semantics validator by at least 2000 bytes and rendered it too costly to execute within the Cardano ledger's present execution-cost limits.

The audit report for Marlowe on Cardano was based on the commit hash [523f3d56](https://github.com/input-output-hk/marlowe-cardano/commit/523f3d56f22bf992ddb0b0c8a52bb7a5a188f9e9). The revisions and mitigations discussed here apply to [ac65eba1](https://github.com/input-output-hk/marlowe-cardano/commit/) of that repository. The appendix to this report list the differences between pre- and post-audit validator code for Marlowe on Cardano.

The medium-severity concerns in the audit report center around missing tests or name shadowing. These have all been remedied either by code changes (in the case of the shadowing) or by the implementation of a significant augmentation of the property-based test suite. The audit's low-severity concerns generally relate to insufficiently detailed text in the Marlowe Cardano specification, the need for more elaborate comments in the code, naming of variable bindings, or typographical errors. Nearly all of these finds have been addressed, with only the exception of a few cosmetic recommendations that were not adopted.

@@

Although this `Eq` instance is not used by the Marlowe validators (so it does not have implications for validator security), the equality test has been fixed because it is a liability for off-chain code and for potential future versions of validators that might perform such an equality test. Commit [84d65a70](https://github.com/input-output-hk/marlowe-cardano/commit/84d65a70c29320fab501ad759bbc71855d1b63fc) fixes this and commit [04805a39](https://github.com/input-output-hk/marlowe-cardano/commit/04805a393c0815fe0f2a1a6d7f76d9867ffe0c14) hardens all of the other `Eq` instances to prevent similar problems re-arising if sum types are augmented in the future.

### 2.1.4 Inaccurate formulation of Money preservation *(Severity: High)*

> **File `specification-v3-rc1.pdf`, Section `3.1 Money preservation`, page *29***
>
> As the property stands, it is permitted to make deposits in one currency and return payments in a different currency. As long as the sums of the amounts match, the equality is satisfied. Yet it is unlikely that the participants of the contract would agree that money has been preserved.
>
> Money preservation is a property stated with an equality. The left hand side is the sum of the deposits done by a list of transactions. The right hand side of the equality is the sum of all the payments done in the same list of transactions. Each sum, in turn, is represented as a single integer which aggregates the amounts of the various payments and deposits, irrespective of what currencies correspond to these amounts.

To address this, a new `MultiAssets.thy` Isabelle theory was created. This theory defines a new type that ties together the number of tokens and the type of token, and it implements some type classes to be able to work with normal addition and subtraction. Also, the `MoneyPreservation` was refactored into `AssetsPreservation`, addressing issue 2.1.4 among others.

These changes can be found in [PR-161](https://github.com/input-output-hk/marlowe/pull/161)

@@

### 2.1.5 Insufficient documentation of Money preservation *(Severity: Medium)*

> **File `specification-v3-rc1.pdf`, Section `3.1 Money preservation`, page *29***
>
> Money preservation is formulated in terms of functions that are not discussed in the specification. It is necessary to explain the meaning of these functions in sufficient detail so readers can understand the property.

The theory was completely refactored in [PR-161](https://github.com/input-output-hk/marlowe/pull/161), adding the necessary documentation.

@@

### 2.1.6 Missing description of Merkleization *(Severity: High)*

> **File `specification-v3-rc1.pdf`, `Merkleization`**
>
> There is no property about merkleization, but merkleization is implemented in the Cardano integration.
>
> Some relevant properties could be:
>
> 1.  The merkleized contract produces the same payments as the analogous regular contract.
> 2.  If a merkleized case input is applied successfully, it implies that the contract hash in the input corresponds to the continuation of the contract.
> 3.  Merkleizing and unmerkleizing a contract gives back the original contract.

We did not mention the merkleization in the specification document as it was considered an implementation detail. This does lessen the applicability of the Isabelle proofs to the different implementations (Marlowe on Cardano, Marlowe on PureScript, etc). To check that an implementation is compliant to the specification we use the `spec-test` tool, which uses property based tests to check that a custom implementation behaves equally to the Isabelle implementation. Without the proper formalization, the tool can only test for the non-merkleized cases. In practice this gives us confidence that the merkleized interpreter works the same for non-merkleized cases. To test the correctness of merkleized cases, the marlowe-cardano repository has specific tests.

The Merkleization formalization will be introduced in a future version of the specification, the initial formulation started in [PR-171](https://github.com/input-output-hk/marlowe/pull/171).

@@

### 2.1.7 Positive balances are not checked for the output state *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 13`**
>
> *Positive balances* are only checked for the input, not for the output Marlowe state. If the semantics are flawed, a transaction can produce an unspendable output that does not satisfy this constraint.
>
> If such a transaction is accepted, no further evaluation will be possible since all subsequent transactions will be rejected due to the very same Constraint 13. This is an hypothetical attack vector, where a malicious actor could send a transaction to block a contract.

The Plutus code for the Marlowe semantics validator had been implemented with the optimistic assumption that semantics would not be flawed and that the additional execution cost of checking the output state was not warranted. However, as the audit report points out, such a flaw would result in the contract being forever blocked from further execution. Adding a check on the final output's validity now prevents a contract from ever reaching a fully blocked state. Even if the semantics or Plutus code were flawed, at least some execution paths in the contract might remain viable, so that the contract could eventually terminate and would not permanently lock funds.

@@

Again as previously, the relevant commits are [0a890845](https://github.com/input-output-hk/marlowe-cardano/commit/0a890845c44ccfe4df97e765e9b9eb743ce7d580), [8855feae](https://github.com/input-output-hk/marlowe-cardano/commit/8855feae473882b82643db792327423152e45b5c), [26f024e8](https://github.com/input-output-hk/marlowe-cardano/commit/26f024e80cea0dd4ebf8bff0f1a10b97aea87894), and [201c5df9](https://github.com/input-output-hk/marlowe-cardano/commit/201c5df9de29b3b142be5ff7d0a9b56d1b378204) for specification and implementation and [d514bcd0](https://github.com/input-output-hk/marlowe-cardano/commit/d514bcd075732b07e1c6a73e2e3c68afa01acf2c), [7f562545](https://github.com/input-output-hk/marlowe-cardano/commit/7f562545d013dd17472b692b550ffdc7f8a383f3), and [ebab31d9](https://github.com/input-output-hk/marlowe-cardano/commit/ebab31d9ba6dd322d374b34cad3995050a0e476e) for property-based tests.

### 2.1.10 Unchecked ending balance *(Severity: High)*

> **File `marlowe-cardano-specification.md`, `Constraint 5`**
>
> The balance of the starting Marlowe state is checked to match the value in the input. However, the validator does not check that the ending balance matches the value in the output paid to the Marlowe validator. Similarly to Issue […], if there are flaws in the semantics that cause the ending balance to differ from the actual value paid to the validator, this constraint would prevent any transaction from spending the output.
>
> The specification should at least discuss why the check is absent together with the other similar checks that are not implemented (checking that ending accounts have positive balances, checking that the ending Marlowe state is valid).

As the audit report recognizes, this concern is closely related to the three previous concerns. The revised validator's enforcement of the three invariants for the final state ensures that the ending balance of the internal accounts in the state matches the actual output paid to the script.

@@

Commit [a2ff6aa3](https://github.com/input-output-hk/marlowe-cardano/commit/a2ff6aa334efb7b7955408d29cf0e2edf06bca08) annotates the Marlowe semantics validator with reasoning as to the safe use of `AssocMap`. Commits [af380a29](https://github.com/input-output-hk/marlowe-cardano/commit/af380a292c0369bcc77d8584d81ea80a31326b61), [e2277677](https://github.com/input-output-hk/marlowe-cardano/commit/e2277677987f359dcd8754275eccd5b7c9a880a7), [a95a616a](https://github.com/input-output-hk/marlowe-cardano/commit/a95a616aca59bc7159d3ffa92487f8f021820bff), [b65ccfec](https://github.com/input-output-hk/marlowe-cardano/commit/b65ccfec3708484ff52514671e0e227c13f084f3), and [9e068b6a](https://github.com/input-output-hk/marlowe-cardano/commit/9e068b6a60103f2a40604a7eaea57c5f29b59180) implement the property-based tests for the behavior of the Plutus `AssocMap` against the Isabelle `MList`.

### 2.1.13 Missing specification tests *(Severity: Medium)*

> **File `Spec/Marlowe/Semantics/Compute.hs`**
>
> There are no tests for the properties in Section 3 of `specification-v3-rc1.pdf`. Besides checking that there are no translation mistakes, these properties would also help contrasting the assumptions in the Isabelle and the Haskell sides, like the meaning of validity of an association list, which is focused in the previous issue.

Subsequent to the commencement of the audit, a "test oracle" (called `test-spec`) was developed that checks implementations of Marlowe semantics (such as the Plutus validator) against a reference implementation generated directly from the Isabelle specification. That generated Haskell implementation is backed by Isabelle proofs of the correctness of Haskell code generation by Isabelle. The test oracle is now applied to the Marlowe semantics implementation in Plutus that the validator use. The test oracle uses the reference implementation to generate test cases to challenge the Plutus implementation and then it checks the Plutus result against its own result. This oracle provides sufficient coverage to address this audit-reports concern regarding the lack of property-based tests for Section 3 of the Marlowe specification.

@@

## 2.2 Marlowe specification

### 2.2.1 Lack of explanation regarding changing choices *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.4 Choices`, page *10***
>
> Choices can only be changed when evaluating `When` statements. This is something only evident after looking at the implementation of `computeTransaction`. It needs to be discussed when first introducing choices and the `When` contract.

We improved documentation regarding choices in [PR-168](https://github.com/input-output-hk/marlowe/pull/168) and [PR-182](https://github.com/input-output-hk/marlowe/pull/182).

@@

### 2.2.2 Undefined reference *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.7 Contracts`, page *13***
>
> There is an undefined reference.

We fixed this in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).

@@

### 2.2.3 Lack of explanation for necessity of `Environment` type *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.8 State and Environment`, page *14***
>
> An `Environment` type is introduced, but it is unclear why it is needed as it is defined as a synonym for time intervals.

We added more documentation regarding the environment in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).

@@

### 2.2.4 Unclear meaning of execution environment *(Severity: Low)*

> **File `specification-v3-rc1.pdf`, Section `2.1.8 State and Environment`, page *14***
>
> The meaning of the execution environment of the transaction is unclear. This is due to the concept of *transaction* being assumed by the specification and never formally introduced.
>
> The specification reads
>
> > The execution environment of a Marlowe contract simply consists of the (inclusive) time interval within which the transaction is occurring.
>
> One has to infer that evaluating a Marlowe contract is undefined if it does not happen within a transaction, as otherwise the description of the execution environment would not make sense. It would be necessary to establish more explicitly the relationship between the contract evaluation and the notion of transaction.

We added more documentation regarding the execution environment in [PR-182](https://github.com/input-output-hk/marlowe/pull/182).

@@

### 2.2.5 Unexplained interval data types *(Severity: Low)*
